---
title: "A4: Incarceration"
author: "Drew King"
date: "08/04/2023"
output: html_document
---

### Introduction

Mass incarceration in the United States is an alarming trend that is on the rise. The United States has the most incarcerated people per capita in the world. In this report, the average prison incarceration per county since 1990 is analyzed by, race, population, and geographic boundaries (States). The analysis is done in this manner to show how there are disparities in incarceration by geographic region and ethnicities.

### Summary Information

The average total change in Incarceration rates from 1990 to 2016 by type of incarceration and race was computed and shown in the following table. This was done to show which racial groups are going into which carceral system. There was not enough data on AAPI populations to include them. The BIPOC variable is an aggregation of black and indigenous people who were incarcerated. Data on Latinx populations was not included due to time constraints and specification requirements of this report. Lastly, the total rate of increase in prison incarceration from 1970 to 2016 was calculated to show the scope of the issue of rising incarceration in the United States, this data is shown below the table.

``````{r, echo = FALSE, message = FALSE, warning = FALSE, code = readLines("summary.R")}
```

### The Dataset

Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)


### Trends Over Time Chart: Incarceration by Race

Incarceration by race since 1990 was visualized in this report to create more understanding about:

* Increases in incarceration by year
* Who is being incarcerated the most
* Where people are being incarcerated

Patterns that emerged are:

* Black and Indigenous People are an overwhelming majority of the United States carceral population.
* Jail rates and prison rates are steadily rising for white people.
* Incarceration of the BIPOC population is cooling down since 2008, but still much higher than that of white people.

``````{r, echo = FALSE, message = FALSE, warning = FALSE, code = readLines("chart1_example.R")}
```

### Variable Comparison Chart

Incarceration by population growth was visualized in this report to see if there is a correlation between population growth and incarceration growth from 1990 to 2016. Some technical limitations required that only the top 10 most incarcerated states be shown including:

* Color coding by states makes states indiscernable for more than 10 states.
* Some states have shrinking incarcerated populations, making it hard to fit a trend line.
* A legend for all 50 states and if their prisons are growing does not fit well.

Emerging patterns:

* Texas's incarceration rate is growing faster than any other state and so are its prisons
* Texas aside, there is no trend between population growth and incarceration growth.
* States with high incarceration rates have incarcerated more people than they have had people move to their state since 1990
* The top 5 states with growing incarceration, excluding New York State are located in the south.

``````{r, echo = FALSE, message = FALSE, code = readLines("chart2_example.R")}
```

### Map

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The last chart that you will create and include will show how a variable is distributed geographically. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Your map needs a title
- Your color scale needs a legend with a clear label
- Use a map based coordinate system to set the aspect ratio of your map

``````{r, echo = FALSE, message = FALSE, code = readLines("map_example.R")}
```

